---
title: "Exploring"
output: html_document
---

Load libraries:

```{r}
rm(list = ls())
library(tidyverse)
library(npdr)
```

Read data:

```{r}
train_set <- data.table::fread('data/SubCh2/SubCh2_TrainingData_imputed_combat.csv')

test_set <- data.table::fread('data/SubCh2/SubCh2_TestData_combat.csv')
head(colnames(train_set), 10)
train_covars <- c('Sample_Names', 'Country', 'Asexual.stage..hpi.', 
                  'Kmeans.Grp', 'ClearanceRate')
test_covars <- c('Sample_Names', 'Asexual_Stage', 'Country', 'Isolate',
                 'Timepoint', 'Treatment', 'BioRep')
train_feats <- train_set %>%
  dplyr::select(- train_covars)
test_demo <-  test_set %>%
  dplyr::select(test_covars)
test_feats <- test_set %>%
  dplyr::select(- test_covars)
```

```{r}
plot(train_set$Asexual.stage..hpi., train_set$ClearanceRate)
lm(ClearanceRate ~ Asexual.stage..hpi., data = train_set) %>%
  summary()

ggplot(train_set, aes(x = Asexual.stage..hpi., y = ClearanceRate)) +
  geom_jitter()
table(train_set$ClearanceRate)
```


Quick glance at the data:

```{r}
head(colnames(train_set))
table(train_set$ClearanceRate)

summary(train_feats[,1:5]) # Notes: some NAs
boxplot(t(train_feats[1:5, ])) # check that the genes are quantile normalized

head(colnames(test_set), 20)
tail(colnames(test_set))
str(test_set[, 1:10])
```

What are the standard deviations of each gene in training set?

```{r}
gene_sds <- apply(train_feats, 2, sd)
summary(gene_sds)
hist(gene_sds)
# head(gene_sds)
str(train_feats)
which.min(gene_sds)
boxplot(train_feats[, 1:5])
```

How are the gene expressions associated with Treatment in test set?

```{r}
treat <- as.factor(test_set$Treatment)
table(test_set$Treatment)
gene_treat_func <- function(x){
  glm(treat ~ x, family = 'binomial') %>%
    summary() %>% 
    with(coefficients) %>%
    .[2, "Pr(>|z|)"]
}

system.time(gene_treat_p <- apply(test_feats, 2, gene_treat_func))
hist(gene_treat_p)
```

Save an example dataset with filtering:

```{r}
thres <- 0.6
kept_genes <- names(gene_sds)[gene_sds > thres]
train_filtered <- train_set %>%
  dplyr::select(train_covars, kept_genes)

write.csv(train_filtered, file = 'data/SubCh2/SubCh2_TrainingData_imputed_combat_ex_filtered.csv')
```

Try NPDR:

```{r}
# cc.data <- train_set %>%
#   dplyr::select(ClearanceRate, names(train_feats))
# 
# system.time(
#   npdr.cc.results <- 
#     npdr('ClearanceRate', cc.data, regression.type="binomial", 
#          attr.diff.type = "numeric-abs", nbd.method = "multisurf", 
#          nbd.metric = "manhattan", msurf.sd.frac = .5, 
#          padj.method = "bonferroni", verbose = T) # %>%
#     # mutate(npdr.log10 = -log10(pval.adj))
# )
# 
# npdr_genes <- npdr.cc.results %>%
#   filter(pval.adj < 0.05) %>%
#   pull(att)
# 
# train_npdred <- train_set %>%
#   dplyr::select(train_covars, npdr_genes)
# 
# write.csv(train_npdred, file = 'data/SubCh2/SubCh2_TrainingData_imputed_combat_npdred.csv')
# save(npdr.cc.results, file = 'results/npdr_results.Rdata')
```

Generate average time point:

```{r}
test_ut <- test_set %>%
  filter(Treatment == 'UT')

alpha <- 1/3
test_ut_24 <- test_ut %>%
  filter(Timepoint == '24HR') %>%
  dplyr::select(Isolate, BioRep, colnames(train_feats))
test_ut_6 <- test_ut %>%
  filter(Timepoint == '6HR') %>%
  dplyr::select(Isolate, BioRep, colnames(train_feats))

# test_ut_6[,1] == test_ut_24[,1]
# test_ut_6[,2] == test_ut_24[,2]

test_ut_24_mat <- test_ut_24 %>%
  dplyr::select(- Isolate, - BioRep) %>%
  as.matrix()
test_ut_6_mat <- test_ut_6 %>%
  dplyr::select(- Isolate, - BioRep) %>%
  as.matrix() 
test_ut_avg_time <- (1/3*test_ut_6_mat + 2/3 * test_ut_24_mat) %>%
  data.frame() %>%
  mutate(Isolate = test_ut_6$Isolate,
         BioRep = test_ut_6$BioRep) %>%
  dplyr::select(Isolate, BioRep, everything())

test_x <- test_ut_avg_time %>%
  filter(BioRep == 'BRep2') %>%
  dplyr::select(colnames(train_feats))
boxplot(cbind(t(test_x[1:5, ]), t(train_feats[1:5, ])))

write.csv(test_ut_avg_time, file = 'data/SubCh2/SubCh2_TestData_combat_avg_timepoint.csv')
```

Examine asexual stage (stage of the parasite based on its transcriptional profile):

```{r}
table(test_ut$BioRep)
test_24_1 <- test_ut %>%
  filter(BioRep == 'BRep1' & Timepoint == '24HR') 
test_24_2 <- test_ut %>%
  filter(BioRep == 'BRep2' & Timepoint == '24HR') 
test_6_1 <- test_ut %>%
  filter(BioRep == 'BRep1' & Timepoint == '6HR') 
test_6_2 <- test_ut %>%
  filter(BioRep == 'BRep2' & Timepoint == '6HR') 

boxplot(train_set$Asexual.stage..hpi., 
        test_6_1$Asexual_Stage,
        test_6_2$Asexual_Stage,
        test_6_3$Asexual_Stage,
        test_6_4$Asexual_Stage,
        test_24_1$Asexual_Stage,
        test_24_2$Asexual_Stage)

table(test_6_1$Asexual_Stage)
table(test_6_2$Asexual_Stage)
table(test_6_3$Asexual_Stage)
table(test_24_1$Asexual_Stage)
table(test_24_2$Asexual_Stage)
table(train_set$Asexual.stage..hpi.)
```

What about other biological replicates?

```{r}
test_ut %>%
  filter(!(BioRep %in% c('BRep1', 'BRep2'))) %>%
  .[, 1:4] 
# Looks like isolate 2 and 27 have more than two bioreps.
test_ut %>%
  filter(Isolate == 'isolate_27') %>%
  .[, 1:4] 
test_ut %>%
  filter(Isolate == 'isolate_02') %>%
  .[, 1:4] 
# For both isolates 2 and 27, looks like biorep 2 is sufficient 
# (low enough asexual stage)
```


```{r}
write.csv(test_24_2[,c(2,4)], file = 'data/SubCh2/TestAsexualStage_24_2.csv')
```
Let's look at 20 and 22:

```{r}
test_ut %>%
  filter(Isolate == 'isolate_20') %>%
  .[, 1:4]

test_ut %>%
  filter(Isolate == 'isolate_22') %>%
  .[, 1:4]
```

Overall min of asexual stage
```{r}
min_as_df <- test_ut %>%
  group_by(Isolate) %>%
  summarise(Asexual_Stage = min(Asexual_Stage))

merged_ut <- left_join(min_as_df, test_ut, by = c('Isolate', 'Asexual_Stage'))
merged_ut <- merged_ut[!duplicated(merged_ut$Isolate),] %>%
  dplyr::select(Isolate, BioRep, Timepoint, colnames(train_feats))
merged_short <- merged_ut[, 1:8]
write.csv(merged_ut, file = 'data/SubCh2/SubCh2_TestData_combat_min_as.csv')
```


Within time point:
```{r}
min_as_df <- test_ut %>%
  group_by(Isolate, Timepoint) %>%
  summarise(Asexual_Stage = min(Asexual_Stage))

merged_ut <- left_join(min_as_df, test_ut, by = c('Isolate', 'Asexual_Stage', 'Timepoint'))
merged_ut <- merged_ut[!duplicated(merged_ut[c('Isolate', 'Timepoint')]), ] %>%
  dplyr::select(Isolate, BioRep, Timepoint, Asexual_Stage, colnames(train_feats))
merged_short <- merged_ut[, 1:8]
write.csv(merged_ut, file = 'data/SubCh2/SubCh2_TestData_combat_min_as_24_6.csv')
```


```{r}
library(scales)

sum(train_set$Asexual.stage..hpi. >= 18)
table(train_set$Asexual.stage..hpi.)
summary(test_set$Asexual_Stage)
stage_df <- data.frame(
  Stage = c(train_set$Asexual.stage..hpi., test_set$Asexual_Stage),
           Dataset = c(rep('Training set', nrow(train_set)), rep('Testing set', nrow(test_set))))

stage_plot <- ggplot(stage_df, aes(x = Stage, fill = Dataset)) +
  geom_bar(alpha = 0.5) + 
  scale_y_continuous(trans = log2_trans()) +
  labs(x = 'Developmental stage', y = 'Count (log scale)') +
  theme_bw() + 
  theme(legend.position = c(0.85, 0.8),
        legend.title = element_blank())

stage_plot

ggsave(stage_plot, filename = 'dev-stage.png', height = 3, width = 5)
```

